ExpressionView -- an interactive viewer for modules identified in gene expression data

### 2010-05-26 expressionviewnote/answers-to-reviewers.tex

\documentclass[a4paper]{article}
\usepackage{url}
\usepackage{natbib}
\usepackage{ragged2e}
\usepackage{color}

\newcommand{\Rpackage}[1]{`\texttt{#1}'}
\newcommand{\Rfunction}[1]{`\texttt{#1}'}
\newcommand{\Rclass}[1]{\textsl{#1}}

\newenvironment{myquote}{\begin{quote}\color{blue}}{\end{quote}}

\setlength{\parskip}{12pt}
\setlength{\parindent}{0pt}

\title{ExpressionView -- an interactive viewer for
  modules identified in gene expression data\\[10pt]
  Answers to reviewers}
\author{Andreas L\"uscher, G\'abor Cs\'ardi, \\ Aitana Morton de
  Lachapelle, Zolt\'an Kutalik, \\ and Sven Bergmann}

\begin{document}

\maketitle

\RaggedRight

First of all, we want to thank the reviewers for their thorough work
and useful suggestions. In the following, the reviewers' comments are
typeset in blue, and indented as a block.

\section*{Reviewer 1}

\begin{myquote}
All comments are for minor revisions:

1) BiVisu by Cheng et al shall be cited:
Bioinformatics. 2007;23(17):2342-4. Please briefly elaborate the major
difference(s) of ExpressionView from BiVisu and BicOverlapper.
\end{myquote}

We have added a brief comparison to Bivisu~\citep{cheng07} and
BicOverlapper\citep{santamaria08}.

\begin{myquote}
2) Page 2, line 30: please re-write the sentence "We find that in
almost every situation, the proposed algorithm finds an order that
recovers more than 99\% of the score of the optimal solution and in
most cases, it recovers the correct alignment". 
\end{myquote}

We have rewritten this part.

\begin{myquote}
3) Page 2, line 55, change "any computer" to "a computer".
\end{myquote}

We have corrected this.

\section*{Reviewer 2}

\begin{myquote}
Comments to the Author

This is a very useful tool which users might  benefit if there is
clearly stated in the website, what sort of input format the
application accepts. There does not seem to be a complete and explicit
description of the various types of input format beyond just a
expression matrix.
\end{myquote}

For the matrix reordering, ExpressionView accepts modules in the
format defined in the eisa R package~\citep{csardi10}
(\textit{ISAModules} object) and the
Biclust R package~\citep{kaiser08} (\textit{Biclust} object). These
cover the existing R implementations of biclustering algorithms (to our
knowledge). Moreover, the Biclust package defines a very simple class
for storing biclustering results, so the output of any other algorithm
(implemented in R or not) can be easily transformed into this
format. This is documented in the Biclust package.

For the expression data itself, ExpressionView accepts
\textit{ExpressionSet} objects as the input. This is the standard data
type in BioConductor for storing expression data. E.g. the GCRMA and
other normalization methods all output this format and it is also easy
to create an \textit{ExpressionSet} object from a simple tabular text
file.

Alternatively, ExpressionView can visualize non-expression data,
inputted as a simple matrix, this is explained in the R manual that
comes with the package.

\begin{myquote}
For example, how do we incorporate GO terms and
KEGG pathways into the raw input dataset, or are these also computed
by the application? 
\end{myquote}

These are computed by the R part of the application, using functions
in the eisa R package. There are examples for this in the manual that
comes with the package.

\begin{myquote}
It is not entirely clear to the casual reader,
whether the Flash application takes raw data or the intermediate XML
result from the workflow, which means user may need to use command
line R anyway to generate the data. If it does so, the format of this
needs to be clearly stated for a non-bioinformatics biologist to be
able to load and run the application. 
\end{myquote}

The Flash application takes the intermediate XML file, we made this
clear in Fig.~1. The idea is that the user performing the analysis in
R is usually not the same, as the user browsing the results. So the
(bioinformatician) user generates the data in R, exports it into an
XML file, and the other (life scientist) user explores it using an
ordinary web browser. We have made this distinction clear in the
manuscript.

The file format of the XML file is not relevant in most cases, as the
file is produced by the ExpressionView package, and the user does not
have to produce it by hand. It is only relevant, if the user wants to
import data that was produced by another reordering algorithm, into
the Flash applet. We have created an XML Schema document that precisely
defines the XML data file's format, and also written a document that
discusses the format. These are available on the ExpressionView website.

\begin{myquote}
The tutorials are impressive, but the audio volume could be increased
or spoken by a person with a higher pitch voice. Very difficult to
hear.  
\end{myquote}

We have increased the audio volume in the video tutorials.

\begin{myquote}
ExpressionView source code (for the Flash applet) in winrar
format contains a lot of empty folders and tmp files some of which may
not be necessary. 
\end{myquote}

We have simplified the directory structure of the source code.

\begin{myquote}
The paper could do with at least a few words in the body of the paper
or in the website, to explain what improvement over features of other
existing Expression data biclustering visualisers currently
available. 
\end{myquote}

We have included a brief comparison to Bivisu~\citep{cheng07} and
BiocOverlapper~\citep{santamaria08}, two existing bicluster
visualization software packages.

\begin{myquote}
Moreover, the paper must a least emphasize justification
for why biologists should find this application/algorithm more useful
over other existing applications/algorithms besides just visual
appeal. 
\end{myquote}

We emphasized the advantages of ExpressionView more in the paper:
\begin{itemize}
\item visualization of all the modules and the expression values together.
\item ease of use for the (life scientist) end user, only a web browser is
  needed, no extra programs needed to be installed.
\item the inclusion of meta-data and enrichment data into the same
  environment, with interactive links to corresponding databases.
\end{itemize}

\section*{Reviewer 3}

\begin{myquote}
Comments to the Author

General comments:

Luescher et al. describe software for visualizing modules in gene
expression data using an efficient implementation of matrix
reordering. Such a package is of high practical value for biologists
to analyze and visualize gene expression data, and I therefore
recommend publication as an Applications Note. In particular, I
appreciate the connection to the Bioconductor package, it is well
done.

Specific comments:

Minor:

I would recommend a different name for the package, ExpressionView is
too general.
\end{myquote}

While this is just criticism, we cannot easily change the name of the
package, as it is part of BioConductor now, and the name change would
cause a lot problems for the users and the BioConductor developers.

\begin{myquote}
Biologists should have no hurdles to access and apply the software
\ldots I could not easily install it on my LINUX/Ubuntu computer after
having installed the isa2 package, as recommended by the authors. 
I assume the download and installation of the package will be as
straightforward as the isa2 package by the authors.
\end{myquote}

The ExpressionView package is an official BioConductor package since
April 2010, so its installation should be easy on any platforms
supported by BioConductor (Linux, MS Windows, Apple OSX). We have
updated the installation instructions on the ExpressionView homepage
accordingly.

\begin{myquote}
The authors should consider that some computers might not play the
Flash applets, and should provide short and clear instructions for
installation on their website.
\end{myquote}

We have added a note about this, and a link to the relevant Adobe
website, to the ExpressionView homepage.

\bibliographystyle{natbib}
\bibliography{expressionview}

\end{document}

### 2010-05-26 expressionviewnote/expressionview.tex
EV note: updated.

\vspace*{-7pt}
\section{Introduction}
Biclustering is an unsupervised data analysis method which is
frequently used to explore microarray data. 
Biclustering algorithms process collections of expression profiles to
identify groups of genes co-expressed under some conditions (samples).
We refer to such groups as modules. While
there is a multitude of biclustering software available (for a reviews 
and comparisons see~\cite{madeira04,ihmels04,prelic06}), packages with 
intuitive interfaces that allow for an interactive exploration of the results are
sparse~\citep{santamaria08}. ExpressionView is designed to close this
gap, and facilitate the collaboration between bioinformaticians and
life scientists without consolidated programming
experience. Implemented as an R package, the user can apply all the
powerful microarray analysis tools provided by the
Bioconductor project~\citep{gentleman04} before exporting the data to
a platform independent Adobe Flash applet. This applet
provides an integrated, intuitive and interactive visualization of
the modules together with the underlying gene expression data and
additional information on the genes and conditions.

\vspace*{-7pt}
\section{Package design and workflow}
The ExpressionView package contains two independent parts: An ordering
algorithm implemented in C++ and an Adobe Flash applet written in
ActionScript and Adobe Flex. Fig.~\ref{fig:workflow} schematically
summarises the typical workflow that uses the gene expression data
available as a Bioconductor ExpressionSet and combines it with the
biclustering results to produce an ExpressionView data file that can
be explored with the Flash applet. In the following, we describe the
different steps in more detail.
\begin{figure*}[!tpb]
\centerline{\includegraphics[width=0.7\linewidth]{fig1-crop}}
\caption{How to use ExpressionView: Starting from gene expression data
  in the form of a Bioconductor ExpressionSet, the user first runs a
  biclustering algorithm to determine co-expressed groups of genes and
  the relevant samples. In a second step, the rows and columns of the gene
  expression matrix are rearranged to produce an easily readable
  overview of the results. The last step consists of combining the
  gene expression data and its associated metadata with the results
  from the biclustering and produce an ExpressionView data file that
  can be explored with the interactive Flash applet.}\label{fig:workflow}
\end{figure*}
For users who prefer to install the viewer as a standalone program, we
also provide an Adobe AIR version that can be downloaded from the
\href{http://www.unil.ch/cbg/ExpressionView}{website}.

\subsection{Gene expression data and modules}
ExpressionView is designed to work with gene expression data in the
form of a Bioconductor ExpressionSet. This class provides a
user-friendly way to access the actual gene expression matrix and its
associated metadata. ExpressionView can treat results obtained by the 
Iterative Signature Algorithm~\citep{bergmann03,csardi09} and any of 
the methods available in the Biclust package~\citep{kaiser08}. Since 
the structure of biclustering results is independent of the algorithm, 
an extension to other methods is straightforward.

\subsection{Reordering genes and conditions}
To present the collection of possibly overlapping modules in a visually
appealing form, it is necessary to reorder the rows (conditions) and
columns (genes) of the gene expression matrix in such a way that biclusters form contiguous
rectangles. Since it is in general impossible to find such an
arrangement for more than two mutually overlapping modules, we propose
here an approximate solution that optimizes the arrangement within the
original data, by maximising the total area of the largest contiguous
module subsets. (An alternative would be to repeat rows and columns as
necessary~\citep{grothaus06}, but for many modules this results in
a very large expression matrix.)

@@

To determine the efficiency of this method, we have studied a large
number of perfectly orderable, but initially scrambled, test cases. We
find that in almost every situation, the proposed algorithm finds an
order that recovers more than 99\% of the score of the optimal
solution and in most cases, it recovers the correct
alignment. For random samples, which are more representative for
actual gene expression data, the execution time increases polynomially
with the number of clusters $m$ as ${\mathcal O}(m^\alpha)$, where
$\alpha \in [1.6, 2]$, almost independent of the number of elements
$n$. For a given number of clusters, we find ${\mathcal O}(n^\alpha)$,
with $\alpha \in [2.5, 2.7]$.

\subsection{Export}
Once the optimal order is determined, the program rearranges the gene
expression matrix accordingly and exports all the relevant information
into an XML file. This format has the advantage of being
self-explanatory and extensible. In addition, the files can be read
and edited with any text editor. The actual gene expression values are
rounded to two digit precision and encoded in Base64. The structure of
the XML file is described in more detail on the ExpressionView
\href{http://www.unil.ch/cbg/ExpressionView}{website}. In most cases,
these files are of the order of a few Megabytes and can thus easily be
sent by e-mail.

\subsection{Visualisation}
The core application of this package is the Flash applet that allows
the user to interactively explore the gene expression data and the
modules. The Flash architecture has the advantage of
producing a platform and browser independent applet that can be
launched from any computer with even the most restrictive user
rights. A screenshot is shown in Fig.~\ref{fig:workflow}. The
interface is divided in two parts: On the left-hand side, the user
finds the gene expression data in the common heat map form, on top of
which the modules are overlaid. On the right-hand side, the metadata
associated with the expression data and the results of the
enrichment calculations for GO~\citep{ashburner00} categories and
KEGG~\citep{kanehisa04} pathways are shown. Wherever possible, these elements
are linked to the corresponding databases. The interface essentially
behaves as an image viewer, allowing the user to zoom and pan around
the expression data, getting instant feedback on the selected
item. The currently selected region can be exported to a PDF file (e.g. 
to prepare figures for presentations or articles) or to a
text file.

\vspace*{-7pt}
\section*{Acknowledgement}

@@

\vspace*{-7pt}
\bibliographystyle{natbib}
\bibliography{expressionview}

>>>>

\vspace*{-9pt}
\section{Introduction}
Biclustering is an unsupervised data analysis method which is
frequently used to explore microarray data. Biclustering algorithms
process collections of expression profiles to identify groups of genes
co-expressed under some conditions (samples). We refer to such groups
as modules. While there is a multitude of biclustering software
available (for reviews  and comparisons
see~\cite{madeira04,ihmels04,prelic06}), packages with intuitive
interfaces that allow for an interactive exploration of the results
are sparse.

Two existing approaches are Bivisu~\citep{cheng07}, an interactive
biclustering program, it plots modules individually, thus the
relationship between the overlapping modules is hard to identify; and
BicOverlapper~\citep{santamaria08}, a novel tool for plotting
overlapping modules in an abstract space. Our approach in
ExpressionView is different, as we use the usual gene-sample
space and visualize all modules together, on top of the reordered
expression matrix. The reordering ensures that, the genes and samples
that appear in the same module, are kept together.

\vspace*{-9pt}
\section{Package design and workflow}
With the ExpressionView package, bicluster analysis can be separated
into two parts. The first part involves finding the
modules in the data set with some algorithm, possibly running
enrichment analysis for the modules, and reordering the rows and
columns of the expression matrix according to the modules. This
part is typically done by a bioinformatician. The second part of the
analysis involves the visualization and interactive exploration of the
results. This part is typically done by researchers without extensive
programming knowledge.

ExpressionView has thus two parts. The first part is written in GNU
R~\citep{R} and contains an implementation of the matrix reordering
algorithm. The second part is an interactive visualization tool, in
the form of an Adobe Flash applet, the user of this part only needs a
Flash-enabled web browser.

This dual implementation has the advantage that all the power of the
GNU R environment and the BioConductor~\citep{gentleman04} packages
can be used for the analysis itself, e.g. all organisms that are (and
will be) supported by BioConductor are automatically supported by
ExpressionView. On the other hand, the exploration of the results does
not need any GNU R knowledge, in most cases no extra software needs to
be installed. Please see Fig.~\ref{fig:workflow} for a typical
ExpressionView workflow.

\begin{figure*}
\centering
\includegraphics[width=0.7\linewidth]{fig1}
\caption{
  Workflow of ExpressionView, showing the two parts of the analysis.
  \textbf{(A)} These steps are performed by a bioinformatician,
  using GNU R. Starting from gene expression data in the form of a
  Bioconductor \textit{ExpressionSet}, the first step is finding the
  modules. In a second step, the rows and columns of the gene
  expression matrix are rearranged to produce an easily readable
  overview of the results. The last step consists of combining the
  gene expression data and its associated metadata, possibly including
  results of enrichment analysis, with the results
  from the biclustering and produce an ExpressionView data file.
  \textbf{(B)}~This file can be distributed and finally explored with
  the interactive Flash applet by the end-user. Please see the
  website for details on the data file format.}
\label{fig:workflow}
\end{figure*}

\vspace*{-9pt}
\subsection{Reordering genes and conditions}

% \subsection{Gene expression data and modules}
ExpressionView is designed to work with gene expression data in the
form of a Bioconductor \textit{ExpressionSet}. This class provides a
user-friendly way to access the actual gene expression matrix and its
associated metadata. ExpressionView can treat results obtained by the 
Iterative Signature Algorithm~\citep{bergmann03,csardi10} and any of 
the methods available in the Biclust package~\citep{kaiser08}. Since 
the structure of biclustering results is independent of the algorithm, 
an extension to other methods is straightforward.

To present the collection of possibly overlapping modules in a visually
appealing form, it is necessary to reorder the rows (conditions) and
columns (genes) of the gene expression matrix in such a way that
biclusters form contiguous
rectangles. Since it is in general impossible to find such an
arrangement for more than two mutually overlapping modules, we propose
here an approximate solution that optimizes the arrangement within the
original data, by maximizing the total area of the largest contiguous
module subsets. (An alternative would be to repeat rows and columns as
necessary~\citep{grothaus06}, but for many modules this results in
a very large expression matrix.)

@@

We have studied a large number of perfectly orderable, but initially
scrambled, test cases. We find that the proposed algorithm finds an
order that recovers more than 99\% of the score of the optimal
solution and in most cases, it recovers the correct
alignment. For random samples, which are more representative for
actual gene expression data, the execution time increases polynomially
with the number of clusters $m$ as ${\mathcal O}(m^\alpha)$, where
$\alpha \in [1.6, 2]$, almost independently of the number of elements
$n$. For a given number of clusters, we find ${\mathcal O}(n^\alpha)$,
with $\alpha \in [2.5, 2.7]$.

Once the optimal order is determined, the program rearranges the gene
expression matrix accordingly and exports all the relevant information
into an XML file, that can be placed on a web-server or distributed by
email and then imported by the interactive viewer.

\subsection{Visualisation}

A screenshot of the viewer is shown in Fig.~\ref{fig:workflow}. The
interface is divided in two parts: On the left-hand side, the user
finds the gene expression data in the common heat map form, on top of
which the modules are overlaid. On the right-hand side, the metadata
associated with the expression data and the results of the
enrichment calculations for GO~\citep{ashburner00} categories and
KEGG~\citep{kanehisa04} pathways are shown. Wherever possible, these elements
are linked to the corresponding databases. The interface essentially
behaves as an image viewer, allowing the user to zoom and pan around
the expression data, getting instant feedback on the selected
item.

\vspace*{-9pt}
\section*{Acknowledgement}

@@

\vspace*{-9pt}
\bibliographystyle{natbib}
\bibliography{expressionview}
